Annotation-Based Empirical Performance Tuning Using Orio

### 2009-02-15 doc/papers/ipdps09/integration.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@443 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

Figure~\ref{fig:pluto-orio} outlines the overall structure of the
Pluto-Orio integrated system, which is implemented as a new
optimization module in Orio. Initially, the transformation process
takes two kinds of input: the loop code to be optimized and the
optimization parameters code. The loop code is embedded in the
annotation body block, while the performance parameters code is
written in the module body block. A simple parser implemented inside
the new module parses the performance parameters code to extract their
values. In addition to the loop code to be optimized, the module
parses performance parameters, which include tile sizes, unroll
factors, loop order (permutation), as well as several boolean values
for triggering OpenMP parallelization, scalar replacement, and
auto-vectorization. Using these parameter values, Pluto then performs
polyhedral transformations (e.g., two-level tiling and OpenMP
parallelization) on the input loop code. The resulting generated code
is subsequently passed to the widely-available profiling tool
gprof~\cite{gprof} for hotspot detection. The statistical information
produced by gprof provides accurate locations (line numbers) of the
loop nests where the Pluto-generated code spends most of its execution
time. The identified hotspot loop nests are then decorated with Orio's
performance annotations for complementary syntactic transformations
(e.g., loop permutation, loop unroll/jamming, scalar replacement, and
explicit auto-vectorization). Finally, Orio optimizes the annotated
hotspots as described in Section~\ref{sec:implm}.

>>>>

Figure~\ref{fig:pluto-orio} outlines the overall structure of the
Pluto-Orio integrated system, which is implemented as a new
optimization module in Orio. Initially, the transformation process
takes two kinds of input: the loop code to be optimized and the
optimization parameters code. The loop code is embedded in the
annotation body block, while the performance parameters code is
written in the module body block. A simple parser implemented inside
the new module parses the performance parameters code to extract their
values. The extracted values of the performance parameters include
tile sizes, unroll factors, loop order (permutation), as well as
several boolean values for triggering OpenMP parallelization, scalar
replacement, and auto-vectorization. Using these parameter values,
Pluto then performs polyhedral transformations (e.g., two-level tiling
and OpenMP parallelization) on the input loop code. The resulting
generated code is subsequently passed to the widely-available
profiling tool gprof~\cite{gprof} for hotspot detection. The
statistical information produced by gprof provides accurate locations
(line numbers) of the loop nests where the Pluto-generated code spends
most of its execution time. The identified hotspot loop nests are then
decorated with Orio's performance annotations for complementary
syntactic transformations (e.g., loop permutation, loop
unroll/jamming, scalar replacement, and explicit
auto-vectorization). Finally, Orio optimizes the annotated hotspots as
described in Section~\ref{sec:implm}.

### 2009-02-15 doc/papers/ipdps09/results.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@437 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@439 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@443 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\begin{figure}%[b]
\begin{tabular}{c}
\begin{minipage}[t]{.47\textwidth}
\centering
\vspace{0.2in}
\includegraphics[width=1.0\textwidth]{figures/spmv.eps}  
\end{minipage}
\\
{\footnotesize (a)}
\\
\begin{minipage}[t]{.3\textwidth}
\scriptsize
\begin{verbatim}

@@

\end{verbatim}
\end{minipage}
\\
{\footnotesize (b)}
\\
\end{tabular}

@@

\begin{figure*}%[t] 
\begin{center} 
  \subfigure[Sequential (single core)]{ 
  \includegraphics[width=.45\textwidth]{figures/axpy4_bgp/seq.eps}  
  \label{fig:axpy4-bgp-seq} 
  } 
  \subfigure[Parallel (four cores)]{ 
  \includegraphics[width=.45\textwidth]{figures/axpy4_bgp/par.eps}  
  \label{fig:axpy4-bgp-par} 
  } 
\end{center}
\caption{Performance of AXPY-4 operations on the Blue Gene/P.} 
\label{fig:axpy4-bgp-results} 
\end{figure*}

The performance results shown in Figure~\ref{fig:axpy4-bgp-results} indicate
that the code tuned by Orio consistently outperforms the other three versions
for both the sequential and parallel cases. We observe that even for a simple
algebraic operation, such as the composed AXPY routines, the compiler alone
is unable to yield performance comparable to the empirically tuned
version. Moreover, implementations that rely on calls to multiple tuned
library routines (e.g., Goto BLAS and ESSL) suffer from loss of both spatial
and temporal localities, resulting in inferior memory performance.
%The sequential BLAS-based implementation perform better than the
%sequential single loop implementation. while for the parallel case, the 
%simple loop implementation is more efficient than the BLAS codes.

@@

The SpMV operation computes $\forall_{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
one should use compact data structures for $A$ and try to maximize temporal
reuse of $x$ and $y$. One of the most commonly used data structures for
storing a sparse matrix is compressed sparse row (CSR) storage
~\cite{vuduc-thesis},
illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
packed together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. An implementation of SpMV using CSR storage is
shown in Figure~\ref{fig:spmv}(b).

@@

The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again show
%Blue Gene/P was done for a varied number of nodes (i.e., 1, 8, and 32)
%and the same three different parallel modes as before (i.e., SMP,
%dual, and VN). 
that empirical optimization using Orio produces the best performance for
all cases. 
%outperforming both the hand-tuned and the xlc-optimized codes. 
On this architecture, the performance gap between the xlc-optimized and the
hand-tuned codes is now larger than in the Intel experiment. Similarly, by
exploiting thread-level parallelism, the Orio-optimized code performs better
than the hand-tuned version when the number of threads per process increases
and the number of nodes decreases. For the single-thread cases (VN mode),
thread-level parallelism is not exploited; nevertheless, the Orio version
still performs better than the other two choices.
%This validates the
%effectiveness of Orio in empirically tuning sequential codes.

@@

In the following sections, we refer to the icc-optimized base code as
xs``ICC,'' the Pluto-generated code with L1 tiling and
unroll/jamming as ``Pluto (L1 tiling),'' and the Pluto-generated code with L1
and L2 tilings and unroll/jamming as ``Pluto (L1+L2 tiling).'' The code
tuned by Orio is referred to as ``Pluto+Orio.''

\begin{figure*}%[htb] 
\begin{center} 
  \subfigure[Sequential (T=500)]{ 
  \includegraphics[width=.45\textwidth]{figures/fdtd-2d-cookie/seq.eps} 
  \label{fig:fdtd-2d-cookie-seq} } \subfigure[Parallel (T=500, 
  N=2000)]{ 
  \includegraphics[width=.45\textwidth]{figures/fdtd-2d-cookie/par.eps} 
  \label{fig:fdtd-2d-cookie-par} } 
\end{center}
\caption{2-D FDTD performance on an eight-core Intel Xeon.} 
\label{fig:fdtd-2d-cookie-results} 
\end{figure*}

@@

Figure~\ref{fig:seidel-cookie-seq} contains the sequential performance
results for $T=500$, which show that applying Pluto's polyhedral tiling on
the original code always delivers performance boosts, which range from 126\%
to 142\%. When using Pluto, performing two-level tiling (for both L1 and L2
caches) yields 13\% lower performance than performing only L1 tiling. 
This is also reflected in the best sequential code found by Orio, where
one-level tiling (for L1 cache only) is always performed. The sequential
speedup obtained from tuning the Pluto code with Orio is significant, ranging
from 51\% to 133\%.

The parallel performance results for $T=500$ and $N=4000$ are shown in
Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc failed to
parallelize the code, whereas Pluto was able to parallelize it and Orio
improved the performance of the one-level tiled Pluto code further by up to
128\% (and up to 548\% for the two-level tiled Pluto code). We observed that
tiling over both L1 and L2 can result in worse performance than L1-only
tiling because the number of tiles can be smaller than the number of cores.
%Similar earlier trend also exists
%in these performance results: Pluto parallelizes the code through its
%polyhedral transformations (i.e., skewing and tiling), while the icc
%compiler alone does not parallelize the loop. Empirical tuning via
%Orio creates up to 122\% speedups over the one-level tiled Pluto code,
%and up to 548\% speedups over the two-level tiled Pluto code. Another
%finding is that tiling for both L1 and L2 caches contributes to a poor
%scalability because the number of tiles can be less than the number of
%processors, resulting in underutilization of the available resources.

>>>>

\begin{figure}%[b]
\begin{tabular}{c}
\begin{minipage}[t]{.43\textwidth}
\centering
\vspace{0.2in}
~~~~~~~~\includegraphics[width=1.0\textwidth]{figures/spmv.eps}  
\end{minipage}
\\
(a)
\\
\begin{minipage}[t]{.3\textwidth}
\scriptsize
\begin{verbatim}

@@

\end{verbatim}
\end{minipage}
\\
(b)
\\
\end{tabular}

@@

\begin{figure*}%[t] 
\begin{center} 
  \subfigure[Sequential (single core)]{ 
  \includegraphics[width=.45\textwidth]{figures/axpy4_bgp/seq.eps}  
  \label{fig:axpy4-bgp-seq} 
  } 
  \subfigure[Parallel (four cores)]{ 
  \includegraphics[width=.45\textwidth]{figures/axpy4_bgp/par.eps}  
  \label{fig:axpy4-bgp-par} 
  } 
\end{center}
\caption{Performance of AXPY-4 operations on Blue Gene/P.} 
\label{fig:axpy4-bgp-results} 
\end{figure*}

The performance results shown in Figure~\ref{fig:axpy4-bgp-results} indicate
that the code tuned by Orio consistently outperforms the other three versions
for both the sequential and parallel cases. We observe that even for a simple
algebraic operation, such as the composed AXPY routines, the compiler alone
is unable to yield performance comparable to the empirically tuned
version. Moreover, implementations that rely on calls to multiple tuned
library routines (e.g., Goto BLAS and ESSL) suffer from loss of both spatial
and temporal localities, resulting in inferior memory performance.

%The sequential BLAS-based implementation perform better than the
%sequential single loop implementation. while for the parallel case, the 
%simple loop implementation is more efficient than the BLAS codes.

@@

The SpMV operation computes $\forall{A_{i,j}} \neq 0:y_{i}
\leftarrow y_{i} + A_{i,j} \cdot x_{j}$, where $A$ is a sparse matrix,
and $x$, $y$ are dense vectors. Each element of $A$ is used precisely once,
and element reuse is possible only for $x$ and $y$. Thus, to optimize SpMV,
one should use compact data structures for $A$ and try to maximize temporal
reuse of $x$ and $y$. One of the most commonly used data structures for
storing a sparse matrix is compressed sparse row (CSR) storage
~\cite{vuduc-thesis},
illustrated in Figure~\ref{fig:spmv}(a). Elements in each row of $A$ are
packed together in a dense array, \texttt{Aval}, and a corresponding array of
integers \texttt{Aind} stores the column indices. The \texttt{Aptr} array
designates where each sparse row begins in \texttt{Aval} and
\texttt{Aind}. An implementation of SpMV using CSR storage is
shown in Figure~\ref{fig:spmv}(b).

@@

The Blue Gene/P results in Figure~\ref{fig:ex27-bgp-results1} again
show that empirical optimization using Orio produces the best
performance for all cases. On this architecture, the performance gap
between the xlc-optimized and the hand-tuned codes is now larger than
in the Intel experiment. Similarly, by exploiting thread-level
parallelism, the Orio-optimized code performs better than the
hand-tuned version when the number of threads per process increases
and the number of nodes decreases. For the single-thread cases (VN
mode), thread-level parallelism is not exploited; nevertheless, the
Orio version still performs better than the other two implementations.

%This validates the
%effectiveness of Orio in empirically tuning sequential codes.

@@

In the following sections, we refer to the icc-optimized base code as
``ICC,'' the Pluto-generated code with L1 tiling and
unroll/jamming as ``Pluto (L1 tiling),'' and the Pluto-generated code with L1
and L2 tilings and unroll/jamming as ``Pluto (L1+L2 tiling).'' The code
tuned by Orio is referred to as ``Pluto+Orio''.

\begin{figure*}%[htb] 
\begin{center} 
  \subfigure[Sequential (T=500)]{ 
  \includegraphics[width=.45\textwidth]{figures/fdtd-2d-cookie/seq.eps} 
  \label{fig:fdtd-2d-cookie-seq} } \subfigure[Parallel (T=500, 
  N=2000)]{ 
  \includegraphics[width=.45\textwidth]{figures/fdtd-2d-cookie/par.eps} 
  \label{fig:fdtd-2d-cookie-par} } 
\end{center}
\caption{2-D FDTD performance on eight-core Intel Xeon.} 
\label{fig:fdtd-2d-cookie-results} 
\end{figure*}

@@

Figure~\ref{fig:seidel-cookie-seq} contains the sequential performance
results for $T=500$, which show that applying Pluto's polyhedral
tiling on the original code always delivers performance boosts, which
range from 126\% to 142\%. When using Pluto, performing two-level
tiling (for both L1 and L2 caches) yields 13\% lower performance than
performing only L1 tiling. This is also reflected in the best
sequential code found by Orio, where one-level tiling (for L1 cache
only) is always found to be better than two-level tiling. The
sequential speedup obtained from tuning the Pluto code with Orio is
significant, ranging from 51\% to 133\%.

The parallel performance results for $T=500$ and $N=4000$ are shown in
Figure~\ref{fig:seidel-cookie-par}. Again we observe that icc failed to
parallelize the code, whereas Pluto was able to parallelize it and Orio
improved the performance of the one-level tiled Pluto code further by up to
128\% (and up to 548\% for the two-level tiled Pluto code). We observed that
tiling over both L1 and L2 can result in worse performance than L1-only
tiling because the number of tiles can be less than the number of cores.
%Similar earlier trend also exists
%in these performance results: Pluto parallelizes the code through its
%polyhedral transformations (i.e., skewing and tiling), while the icc
%compiler alone does not parallelize the loop. Empirical tuning via
%Orio creates up to 122\% speedups over the one-level tiled Pluto code,
%and up to 548\% speedups over the two-level tiled Pluto code. Another
%finding is that tiling for both L1 and L2 caches contributes to a poor
%scalability because the number of tiles can be less than the number of
%processors, resulting in underutilization of the available resources.
